---
title: Software Engineering Standards in the R Community
author:
  - name: Oliver Keyes
    affiliation: Wikimedia Foundation
    email:  ironholds@gmail.com
  - name: Jennifer Bryan
    affiliation: University of British Columbia
    email:  jenny@stat.ubc.ca
  - name: David Robinson
    affiliation: Princeton University
    email:  admiral.david@gmail.com
abstract: >
  Oliver's abstract for useR conference goes here.
preamble: >
  % Any extra latex you need in the preamble
output: rticles::rjournal_article
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, cache = TRUE)
```

```{r load_data}
library(practice)
library(plyr)
library(dplyr)
library(tidyr)
library(ggplot2)

data("CRANpractices")

practice_names <- c(
  casing = "Package Name Case",
  license = "License",
  upstream_repo = "Upstream Repository",
  versioning = "Semantic Versioning",
  testing = "Tests",
  roxygen = "Roxygen Documentation",
  changelog = "Change Log",
  vignette_format = "Vignette Format",
  vignette_builder = "Vignette Builder"
)
```

```{r practices_by_year, dependson = "load_data"}
# compute practices per year, for use in time series plots
library(lubridate)

practices_by_year <- CRANpractices %>%
  select(-alphanumeric, -version, -links_to) %>%
  gather(practice, choice, -package, -date) %>%
  mutate(date = as.Date(date)) %>%
  filter(!is.na(date), year(date) > 2000) %>%
  mutate(year = round_date(date, "year")) %>%
  count(practice, choice, year) %>%
  group_by(practice, year) %>%
  mutate(percent = n / sum(n)) %>%
  ungroup() %>%
  mutate(choice = reorder(choice, n, function(x) -mean(x)))
```

```{r plot_practice, dependson = "practices_by_year"}
plot_practice <- function(.data) {
  .data %>%
    mutate(practice = revalue(practice, practice_names)) %>%
    ggplot(aes(year, percent, color = choice)) +
    geom_line() +
    facet_wrap(~ practice, scales = "free") +
    labs(color = "") +
    theme_bw()
}
```


## Introduction

Introductory section which may include references in parentheses
\citep{R}, or cite a reference such as \citet{R} in the text.

## Best Practices

We considered a number of development practices...

* **Tests** Tests allow... [cite testthat].
* **Vignettes** Vignettes provide long-form documentation, such as a tutorial. Developers have the choice of several vignette builders, of which the most popular are the built-in \pkg{Sweave} and the \pkg{knitr} package. The \pkg{knitr} package also offers a choice of format, between LaTeX and R Markdown.
* **roxygen documentation**: \pkg{roxygen}, and more recently \pkg{roxygen2}, make it easy to...

## Practices in CRAN

Some figures and analyses, like Figure \ref{fig:CRAN_vignettes}.

\begin{figure}
```{r practices_overview, dependson = "load_data", fig.height = 15, fig.width = 12, out.height = "6.75in", out.width = "5.4in"}
num_choices <- 15

freq <- CRANpractices %>%
  select(-links_to, -alphanumeric, -version, -date) %>%
  gather(practice, choice, -package) %>%
  count(practice, choice) %>%
  filter(choice != "NA") %>%
  group_by(practice) %>%
  top_n(num_choices, n) %>%
  ungroup() %>%
  mutate(practice = revalue(practice, practice_names),
         choice = reorder(choice, n, function(x) -mean(x)))

ggplot(freq, aes(choice, n)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ practice, scales = "free") +
  xlab("Choice") +
  ylab("Number of packages") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, face = "bold"))
```

```{r num_licenses, dependson = "load_data"}
num_licenses <- length(unique(CRANpractices$license))
```
\caption{Overview of software development practices on CRAN. Note that only the first `r num_choices` choices of license (out of `r num_licenses` used in CRAN) are shown. \label{fig:practices_overview}}
\end{figure}

### Vignettes

```{r vignette_calculations, dependson = "load_data"}
total_packages <- nrow(CRANpractices)
use_vignettes <- sum(CRANpractices$vignette_format != "None")
use_vignettes_percent <- 100 * use_vignettes / total_packages
```

This section will discuss vignettes.

* Only `r use_vignettes` out of `r total_packages` (`r round(use_vignettes_percent, 1)`) packages currently provide any vignettes.
* Figure \ref{fig:vignettes_time} shows the use of vignettes over time. From 2009 onward, vignettes were infrequently found in CRAN, making up less than 5\% of packages. In 2014, their use grew sharply to about 20\%. Since their inception, knitr and R markdown have each grown in popularity as tools for vignette development, though Sweave and LaTeX are still more common overall.

\begin{figure}
```{r vignettes_over_time, dependson = "plot_practice", fig.height = 5, fig.width = 10, out.width = "6in", out.height = "3in"}
g1 <- practices_by_year %>%
  filter(practice == "vignette_format", choice != "None") %>%
  plot_practice()

g2 <- practices_by_year %>%
  filter(practice == "vignette_builder", choice != "None") %>%
  plot_practice()

library(gridExtra)
print(arrangeGrob(g1, g2, nrow = 1))
```
\caption{Use of particular vignette formats over time.\label{fig:vignettes_time}}
\end{figure}

### Testing

* Most packages don't have tests
* Tests over time (graph)
* Some things that will require new data:
  * Of those that have "Other", what do those tests look like?
  * How many tests do testthat and RUnits have?

### Other practices

\begin{figure}
```{r other_practices_time, dependson = "plot_practice"}
practices_by_year %>%
  filter(practice %in% c("upstream_repo", "versioning", "changelog", "roxygen"),
         !(choice %in% c("FALSE", "None/Other", "Email")),
         !is.na(choice)) %>%
  plot_practice()
```
\caption{Use of semantic versioning etc over time. \label{fig:other_practices_time}}
\end{figure}

* Go over advantages and history of
  * Semantic versioning
  * Upstream repository
  * Change log
* Figure \ref{fig:other_practices_time}, practices are becoming more common. GitHub in particular.

### Conclusion

* Most packages in CRAN don't take advantage of the newest methods and practices in package development. However, that proportion is rapidly increasing in packages published or updated in the last two years.

\bibliography{RJreferences}
